Positivity-Preserving Adaptive Runge--Kutta Methods

### 2020-05-11 nusslein_vsrp/latex/overview.tex
Changed dt of Advection-decay equation
The original dt was set to dt = 0.02 = 2*dx which was too high to expect stability. I have changed it to dt = 0.015 = 1.5*dx which is inside the stability interval

A test problem similar to \cite{shampine_non-negative_2005} is the PDE
\begin{equation}
\label{eq:advection-decay}
\begin{aligned}
  \frac{\partial u(t,x)}{\partial t}
  &=
  -a \frac{\partial u(t,x)}{\partial x} - K u(t,x),
  && x \in (0, 1), t \in (0,1),
  \\
  u(t,0) &= 1,
  && t \in (0,1],
  \\
  u(0,x) &= 0,
  && x \in [0,1],
\end{aligned}
\end{equation}
which consists of an advection part and an exponential decay.
The numerical approximation uses the method of lines and a first order
upwind finite difference semidiscretization with $N = 100$ points.
This leads to the positivity preserving ODE
\begin{equation}
\frac{\mathrm d}{\mathrm d t} u_i = \frac{a}{\Delta x} \left( u_{i-1} - u_i \right) - K u_{i-1}.
\end{equation}
The parameters are set to $a=1$ and $K=1$.
We use the Dormand--Prince RK5 method and adapt the weights using the free adaptation.
In Figure\,\ref{fig:sol_AdDe} the results for $\Delta t = 0.02$ are plotted for different time steps. We can see that the solution approaches an exponential function with $t \rightarrow \infty$.
In Figure\,\ref{fig:weights_AdDe} the used weights are plotted. For $t\leq 0.6$ the weights are altered. For $t > 0.6$ the original weights lead to a positive solution.
\begin{figure}
\centering
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=1\textwidth]{plots/Advection_Decay.pdf}
\caption{Numerical solution at different times.}
\label{fig:sol_AdDe}
\end{subfigure}%
~
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=1\textwidth]{plots/b_Advection_Decay.pdf}\\
\caption{Adaptation of the weights.}
\label{fig:weights_AdDe}
\end{subfigure}
\caption{Numerical results for the advection decay problem \eqref{eq:advection-decay}.}
\end{figure}

>>>>

A test problem similar to \cite{shampine_non-negative_2005} is the PDE
\begin{equation}
\label{eq:advection-decay}
\begin{aligned}
  \frac{\partial u(t,x)}{\partial t}
  &=
  -a \frac{\partial u(t,x)}{\partial x} - K u(t,x),
  && x \in (0, 1), t \in (0,1),
  \\
  u(t,0) &= 1,
  && t \in (0,1],
  \\
  u(0,x) &= 0,
  && x \in [0,1],
\end{aligned}
\end{equation}
which consists of an advection part and an exponential decay.
The numerical approximation uses the method of lines and a first order
upwind finite difference semidiscretization with $N = 100$ points.
This leads to the positivity preserving ODE
\begin{equation}
\frac{\mathrm d}{\mathrm d t} u_i = \frac{a}{\Delta x} \left( u_{i-1} - u_i \right) - K u_{i-1}.
\end{equation}
The parameters are set to $a=1$ and $K=1$.
We use the Dormand--Prince RK5 method and adapt the weights using the free adaptation.
In Figure\,\ref{fig:sol_AdDe} the results for $\Delta t = 0.015$ are plotted for different time steps. We can see that the solution approaches an exponential function with $t \rightarrow \infty$.
In Figure\,\ref{fig:weights_AdDe} the used weights are plotted. For $t\leq 0.375$ the weights are altered. For $t > 0.375$ the original weights lead to a positive solution.
\begin{figure}
\centering
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=1\textwidth]{plots/Advection_Decay.pdf}
\caption{Numerical solution at different times.}
\label{fig:sol_AdDe}
\end{subfigure}%
~
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=1\textwidth]{plots/b_Advection_Decay.pdf}\\
\caption{Adaptation of the weights.}
\label{fig:weights_AdDe}
\end{subfigure}
\caption{Numerical results for the advection decay problem \eqref{eq:advection-decay}.}
\end{figure}
